Prevalence of *Haemoproteus plataleae* in American White Ibis from Southern Florida

### 2021-11-11 products/manuscript/Manuscript.Rmd
Finished Statistical Analysis Part 4 11/11/21
Finished the statistical analysis for the project, organization and manuscript writing left.

---
title: "MADA Project - Ibis Blood Parasites from Florida"
subtitle: ""
author: Ryan Grunert
date: "`r Sys.Date()`"
site: bookdown::bookdown_site
documentclass: article
output:
  bookdown::word_document2: 
    toc: false
    number_sections: true
  bookdown::html_document2: 
    toc: false
bibliography: ../references.bib
csl: ../apa.csl
---
```{r}
#Loading required packages #for data processing
library(here) #to set paths
library(tidyverse) #all required data manipulation packages
library(xlsx) #to read in excel spreadsheets
library(janitor) #To assist in cleaning up dates and tidy data
library(lubridate) #To extract times from raw data
```

@@

## General Background Information
The data I will be using for this project is American White Ibis (*Eudocimus albus*) blood parasite data that was collected from 2010-2017. The parasite the data focuses on is the protozoan *Haemoproteus*, which causes a chronic blood infection in White Ibises and many other birds. *Haemoproteus* is a vector-borne disease, it is transmitted by *Culicoides* biting midges when they feed on the birds and causes long-lasting parasitemia. Adult healthy birds tolerate haemosporidian infections well, but young birds and birds outside their normal range that are naive to the pathogen are more likely to develop clinical disease. Haemosporidian infections usually cause long-lasting chronic infections in natural hosts, and can cause reduced reproductive success, host fitness, and increased stress and disease susceptibility.

Here is a link to a previously published study concerning this data -> https://www.sciencedirect.com/science/article/pii/S221322441730072X

## Description of data and data source
The data was collected in a collaboration between Dr. Michael Yabsley's lab (of which I am a part) and Dr. Sonia Hernandez's lab here at UGA. I received the raw data from Dr. Yabsley in an agreement to analyze the data over the course of the semester in exchange for using the data for the MADA project.

The data is organized in excel spreadsheets. The two main spreadsheets (Ibis_Blood_ParasiteData_2017 and IbisFieldDataOct19_2017) hold the majority of the data, and two supplemental spreadsheets from other collaborators that were or currently working on the project contain additional variables that will be combined with the main dataset in R for additional analyses. Each observation in the data is an individual White Ibis that was captured and sampled, and each variable is a different descriptor of either the Ibis itself, or the site/environment it was captured in.

Two datasheets have overlapping information (Ibis_Blood_ParasiteData_2017 and IbisFieldDataOct19_2017), but both will be combined to have a complete dataset for analysis. The first contains parasitemia data, while the second contains field data. The 100IbisID_UrbanGradient datasheet will be used to assist with the analysis concerning *Haemoproteus* presence and habitat type. The SalmonellaFieldSerotypes datasheet may be used to compare the presence of both *Haemoproteus* and *Salmonella* in the Ibis, although that may not be needed considering the difference in biology and epidemiology of the two pathogens.

```{r}
#Setting the path to the Ibis blood parasite spreadsheet
data_locationIbisB <- here::here("raw_data","Ibis_Blood_ParasiteData_2017.xlsx")

#Loads the raw Ibis blood data excel sheet from 2015-2017
rdIbisBlood15_17 <- read.xlsx(data_locationIbisB, sheetName = "2015-2017")

#Loads the raw Ibis blood data excel sheet from 2010-2014
rdIbisBlood10_14 <- read.xlsx(data_locationIbisB, sheetName = "2010-2014")

str(rdIbisBlood15_17)
str(rdIbisBlood10_14)

```

The main variables I'm interested in for these two sheets above are "Haemoproteus.parasitemia.", and the first "X" column is the sample/bird ID. The 2010-14 set is missing more data, but characteristics of the birds from that group may be used.

```{r}
#Setting the path to the Ibis field data spreadsheet
data_locationIbisF <- here::here("raw_data","IbisFieldDataOct19_2017.xlsx")

#Loads the raw Ibis field data excel sheet
rdIbisField <- read.xlsx(data_locationIbisF, sheetName = "All capture data")

str(rdIbisField)
```

This dataset is the most organized sheet, but columns of interest will be extracted from this sheet as well as the others to create new clean data frames for analysis.

```{r}
#Setting the path to the first Ibis supplemental excel sheet (Urban Gradients)
data_locationIbisUr <- here::here("raw_data","100IbisID_UrbanGradient.xlsx")

#Loads the raw Ibis Urban Gradient supplemental data sheet
rdUrbanIbis <- read.xlsx(data_locationIbisUr, sheetName = "Samples by Date")

str(rdUrbanIbis)

```

These samples from this datasheet will be used in conjunction with the two main datasheets in order to create an "urbanization gradient" for the samples that are located in either rural or urban areas. The gradient will most likely range from least ("1", rural) to most ("4", urban) urban for example, but further work and planning needs to be done first in order to figure out the specifics of the analysis.

```{r}
#Setting the path to the second Ibis supplemental excel sheet (Salmonella)
data_locationIbisSal <- here::here("raw_data","SalmonellaFieldSerotypesAug27_2018.xlsx")

#Loads the raw Ibis Urban Gradient supplemental data sheet
SalIbis <- read.xlsx(data_locationIbisSal, sheetName = "ALL Positives")

str(SalIbis)
```

This sheet details the Ibises that tested positive for Salmonella, but this data may not be used because of the differences in pathogenesis and epidemiology of the two diseases.

## Questions/Hypotheses to be addressed
The questions I plan to address are whether the amounts of *Haemoproteus* parasitemia in the various Ibises are significantly different when focusing on age, sex, season, location, or habitat type (Urban and Rural). Does the amount of parasitemia in the birds decrease, increase or remain the same over time as they get older? Is parasitemia significantly higher in birds sampled in urban or rural environments? These are the main variables I will be looking at for this project, but I may analyze others as the project continues. Not all of the variables were collected for every Ibis sampled, and the data still needs cleaning to accurately detail the sample size for each analysis.

Is there a difference in HaeParasit between Ibis of different ages?
- Dependent variable (outcome): HaeParasit
- Independent variable (predictor): Age

 Is there a difference in HaeParasit between Ibis of different sex?
- Dependent variable (outcome): HaeParasit
- Independent variable (predictor): Sex

Is there a different in HaeParasit in Ibis sampled at different sites?
 - Dependent variable (outcome): HaeParasit
 - Independent variable (predictor): Site

 Is there a difference in HaeParasit between Ibis sampled in different habitat types?
 - Dependent variable (outcome): HaeParasit
 - Independent variable (predictor): HabType

 Is there a difference in HaeParasit between Ibis sampled at different times?
 - Dependent variable (outcome): HaeParasit
 - Independent variables (predictors): Date, Season

## Expected Analyses
I think that I will have to conduct multiple ANOVA's during the analysis process. Linear and logistic regression models apply here too for some variables.I would like to explore the possibility of using the location data for each sample to conduct a geographical analysis with a map as the output. I will be continuing to contact those involved with the study as well to see what analyses they want to be conducted and what questions they want answered.

## Data Cleaning
The data has already been imported, so now we move onto the cleaning stage. The following code will clean, process, and save the raw Ibis blood datasets from 2010-2014 and 2015-2017. This section is contained in the "ProcessingScriptIbisBloodData.R" file.
```{r}
##############################Cleaning the rdIbisBlood10_14 data###########################
IbisBlood10_14 <- rename(rdIbisBlood10_14, ID = X.) #renames the ID column

#Extracts just the columns of interest
IbisBlood10_14 <- select(IbisBlood10_14, ID, Site.ID, Date, PCR.sex, Age..adult.juvenile., 
                         Total.RBC, Haemoproteus.parasitemia.)

#Renames the columns of interest
IbisBlood10_14 <- rename(IbisBlood10_14, Site = Site.ID, Sex = PCR.sex, Age = Age..adult.juvenile.,
                         TotalRBC = Total.RBC, HaeParasit = Haemoproteus.parasitemia.)

#Change "adult" and "juvenile" in the Age column to "A" and "J"
IbisBlood10_14$Age[IbisBlood10_14$Age == "adult"] <- "A"
IbisBlood10_14$Age[IbisBlood10_14$Age == "Adult"] <- "A"
IbisBlood10_14$Age[IbisBlood10_14$Age == "juvenile"] <- "J"
IbisBlood10_14$Age[IbisBlood10_14$Age == "SA"] <- "A"
#Now the Ibises that had age recorded are referred to as "A" for adult or "J" for juvenile

#Changing the DIV/0 values in HaeParasit to NA and change the column to numerical values
IbisBlood10_14$HaeParasit <- as.numeric(IbisBlood10_14$HaeParasit)  # Divide-by-zero errors are now NA

#Reformatting the date values in the Date column
IbisBlood10_14$Date <- excel_numeric_to_date(IbisBlood10_14$Date)
#After this step, the values that just state the year are labelled with "1905", may need to go back or exclude from analysis

#Abbreviating the site names to make analysis easier
IbisBlood10_14$Site[IbisBlood10_14$Site == "Royal Palms"] <- "RP"
IbisBlood10_14$Site[IbisBlood10_14$Site == "Everglades"] <- "E"
IbisBlood10_14$Site[IbisBlood10_14$Site == "Dreher Park"] <- "DRP"
IbisBlood10_14$Site[IbisBlood10_14$Site == "Indian Creek"] <- "ICP"
IbisBlood10_14$Site[IbisBlood10_14$Site == "Indian Creek "] <- "ICP"
IbisBlood10_14$Site[IbisBlood10_14$Site == "Indian Creek Park"] <- "ICP"
IbisBlood10_14$Site[IbisBlood10_14$Site == "Juno Beach"] <- "JB"
IbisBlood10_14$Site[IbisBlood10_14$Site == "Lion Country Safari"] <- "LCS"
IbisBlood10_14$Site[IbisBlood10_14$Site == "Lake Worth"] <- "LW"
IbisBlood10_14$Site[IbisBlood10_14$Site == "Prosperity Oaks"] <- "PO"
IbisBlood10_14$Site[IbisBlood10_14$Site == "Promenade Plaza"] <- "PP"
IbisBlood10_14$Site[IbisBlood10_14$Site == "San Mateo"] <- "SM"
IbisBlood10_14$Site[IbisBlood10_14$Site == "Solid Waste Authority"] <- "SWA"
IbisBlood10_14$Site[IbisBlood10_14$Site == "West Palm Beach Zoo"] <- "WPBZ"

#Reformatting the mislabeled dates. These are set to Jan 1st of the year they were
#recorded on the data sheet, these entries only had the year stated, not month and day.
IbisBlood10_14$Date[IbisBlood10_14$Date == "1905-07-02"] <- "2010-01-01"
IbisBlood10_14$Date[IbisBlood10_14$Date == "1905-07-03"] <- "2011-01-01"
IbisBlood10_14$Date[IbisBlood10_14$Date == "1905-07-04"] <- "2012-01-01"

#The bottom of the dataset has a couple rows of just missing data for all variables,
#This code filters the dataset so that if any rows are present that has at least
#one variable be NOT NA, R keeps it
IbisBlood10_14 <- filter(IbisBlood10_14, if_any(everything(), ~ !is.na(.)))

#This completes the cleaning phase for IbisBlood10_14
#------------------------------------------------------------------------------------#

############################Cleaning the rdIbisBlood15_17 Data#########################

IbisBlood15_17<- rename(rdIbisBlood15_17, ID= X.) #renames the ID column

####Extracting only the columns relevant for analysis
IbisBlood15_17 <- select(IbisBlood15_17,ID,Site.Name,Date,PCR.sex,Age,Season,Habitat.type,
                         Total.RBC,X..Haemoproteus,Haemoproteus.parasitemia.,F1,F2,F3,
                         Avg.RBC.FOV,Ibis..,Total.mass..g.,Mass.bag..g.,Mass.bird..g.,Body.condition.score..1.5.,
                         Ectoparasite.score..1.5.,Culmen.length..mm.,Wing.chord.length..mm.,
                         Tarsus.length..mm.,Tarsus.width..mm.)

####Renaming the columns 
IbisBlood15_17 <- rename(IbisBlood15_17,Site = Site.Name,Sex = PCR.sex,HabType = Habitat.type,TotalRBC = Total.RBC,NumHae = X..Haemoproteus,
                         HaeParasit = Haemoproteus.parasitemia.,AvgRBC = Avg.RBC.FOV,
                         IbisNum = Ibis..,TotalMassG = Total.mass..g.,BagMassG = Mass.bag..g.,
                         BirdMassG = Mass.bird..g.,BodyCondScore = Body.condition.score..1.5.,
                         EctoParasitScore = Ectoparasite.score..1.5.,CulmenLmm = Culmen.length..mm.,
                         WingChordLmm = Wing.chord.length..mm.,TarsusLmm = Tarsus.length..mm.,
                         TarsusWmm = Tarsus.width..mm.)

####Site column
#Ibis that have sites matched with the 10_14 dataset will have the same labels
IbisBlood15_17$Site[IbisBlood15_17$Site == "Juno Beach"] <- "JB"
IbisBlood15_17$Site[IbisBlood15_17$Site == "Indian Creek Park"] <- "ICP"
IbisBlood15_17$Site[IbisBlood15_17$Site == "Dubois park"] <- "DUP"
IbisBlood15_17$Site[IbisBlood15_17$Site == "Dreher Park"] <- "DRP"
IbisBlood15_17$Site[IbisBlood15_17$Site == "Lion Country Safari"] <- "LCS"
IbisBlood15_17$Site[IbisBlood15_17$Site == "Dubois Park"] <- "DUP"
IbisBlood15_17$Site[IbisBlood15_17$Site == "Loxahatchee Wildlife Refuge"] <- "LOXWR"
IbisBlood15_17$Site[IbisBlood15_17$Site == "J.W. Corbett Wildlife Management Area"] <- "JWC"
IbisBlood15_17$Site[IbisBlood15_17$Site == "Gaines Park"] <- "GP"
IbisBlood15_17$Site[IbisBlood15_17$Site == "Solid Waste Authority"] <- "SWA"
IbisBlood15_17$Site[IbisBlood15_17$Site == "Fisheating creek"] <- "FC"
IbisBlood15_17$Site[IbisBlood15_17$Site == "Kitching Creek"] <- "KC"
IbisBlood15_17$Site[IbisBlood15_17$Site == "Loxahatchee NE"] <- "LOXNE"
IbisBlood15_17$Site[IbisBlood15_17$Site == "DuBois Park"] <- "DUP"
IbisBlood15_17$Site[IbisBlood15_17$Site == "TetraTech"] <- "TT"
IbisBlood15_17$Site[IbisBlood15_17$Site == "Green Cay"] <- "GC"
IbisBlood15_17$Site[IbisBlood15_17$Site == "Dreher park"] <- "DRP"
IbisBlood15_17$Site[IbisBlood15_17$Site == "CROW Wildlife Rehabilitation Center"] <- "CROW"
IbisBlood15_17$Site[IbisBlood15_17$Site == "Lake Worth"] <- "LW"
IbisBlood15_17$Site[IbisBlood15_17$Site == "Royal Palm"] <- "RP"
IbisBlood15_17$Site[IbisBlood15_17$Site == "JW Corbett"] <- "JWC"
IbisBlood15_17$Site[IbisBlood15_17$Site == "Busch Wildlife Sanctuary"] <- "BWS"
IbisBlood15_17$Site[IbisBlood15_17$Site == "Loxahatchee Slough"] <- "LOXS"
IbisBlood15_17$Site[IbisBlood15_17$Site == "Indian Creek"] <- "ICP"
IbisBlood15_17$Site[IbisBlood15_17$Site == "Cat House (Bonnie's, Richard Lane)"] <- "CAT"
IbisBlood15_17$Site[IbisBlood15_17$Site == "Palm Beach Zoo"] <- "WPBZ"

####Date column
#Changing the date column to numeric values to allow date extraction
IbisBlood15_17$Date <- as.numeric(IbisBlood15_17$Date) 
##Reformatting the date values in the Date column
IbisBlood15_17$Date <- excel_numeric_to_date(IbisBlood15_17$Date)

####Sex column
IbisBlood15_17$Sex[IbisBlood15_17$Sex == "not bled"] <- NA
#Two values have missing data because the birds were not bled to determine sex
IbisBlood15_17$Sex[IbisBlood15_17$Sex == "M?"] <- NA
IbisBlood15_17$Sex[IbisBlood15_17$Sex == "F?"] <- NA
IbisBlood15_17$Sex[IbisBlood15_17$Sex == "F "] <- "F"
#Two values have question marks for unsure sex determination, removing those values

####Age column
#When an ibis reaches 4 years old, their plumage becomes all white. Any Ibis
#that was sampled with brown plumage is considered a juvenile. Since Haemoproteus is
#a chronic disease that mostly likely accumulates over time, long-lived birds should
#theoretically have more parasitemia. Could not be true though and may come back to
#look at the individual years in the juvenile birds
IbisBlood15_17$Age[IbisBlood15_17$Age == "Juvenile"] <- "J"
IbisBlood15_17$Age[IbisBlood15_17$Age == "Adult"] <- "A"
IbisBlood15_17$Age[IbisBlood15_17$Age == "2nd/3rd year"] <- "J"
IbisBlood15_17$Age[IbisBlood15_17$Age == "3rd year"] <- "J"
IbisBlood15_17$Age[IbisBlood15_17$Age == "1st year"] <- "J"
IbisBlood15_17$Age[IbisBlood15_17$Age == "adult"] <- "A"
IbisBlood15_17$Age[IbisBlood15_17$Age == "2nd year"] <- "J"
IbisBlood15_17$Age[IbisBlood15_17$Age == "NA"] <- NA
IbisBlood15_17$Age[IbisBlood15_17$Age == "2nd Year"] <- "J"
IbisBlood15_17$Age[IbisBlood15_17$Age == "Late third year"] <- "J"
IbisBlood15_17$Age[IbisBlood15_17$Age == "ADULT"] <- "A"
IbisBlood15_17$Age[IbisBlood15_17$Age == "Adult bc"] <- "A" #Not sure what bc stands for

#Season and HabType are fine the way they are

####TotalRBC
IbisBlood15_17$TotalRBC <- as.numeric(IbisBlood15_17$TotalRBC)
#DIV/0 errors are now NA in this column

#NumHae column is fine

####HaeParasit
IbisBlood15_17$HaeParasit <- as.numeric(IbisBlood15_17$HaeParasit)

####-999 values
IbisBlood15_17[IbisBlood15_17 == -999] <- NA

####Removing the last couple empty rows
IbisBlood15_17 <- IbisBlood15_17[-c(401, 402, 403, 404, 405, 406),]

#This completes the cleaning phase for IbisBlood15_17
#------------------------------------------------------------------------------------#

###############################Saving the Cleaned Datasets###########################

#Saving the Ibis Blood 10-14 dataset as an RDS in the processed data folder
save_data_location <- here::here("processed_data", "processedIbisBlood10_14.rds")
saveRDS(IbisBlood10_14, file = save_data_location)

#Saving the Ibis Blood 15-17 dataset as an RDS in the processed data folder
save_data_location2 <- here::here("processed_data", "processedIbisBlood15_17.rds")
saveRDS(IbisBlood15_17, file = save_data_location2)

```

This next section will clean, process, and save the raw Ibis Field dataset. This section is contained in the "ProcessingScriptIbisFieldData.R" file.

```{r}
######################### #Cleaning the rdIbisField Data###############################

IbisField <- rename(rdIbisField, ID = ID..NSFSITE...) #renames the ID column

#Selecting the columns of interest to this analysis from the datasheet
IbisField <- select(IbisField,ID,Site.Name,Latitude,Longitude,Date,PCR.Sex,Age,Season,
                    Habitat.type,Ibis.number,Mass.bird..g.,Body.condition.score..1.5.,
                    Ectoparasite.score..1.5.,Culmen.length..mm.,Wing.chord.length..mm.,
                    Tarsus.length..mm.,Tarsus.width..mm.,Habituation.score.of.flock..1.5.)

#Renaming the columns of interest, also matching the other cleaned datasets
IbisField <- rename(IbisField,Site = Site.Name,Lat = Latitude,Long = Longitude,Sex = PCR.Sex,
                    HabType = Habitat.type,IbisNum = Ibis.number,BirdMassG = Mass.bird..g.,
                    BodyCondScore = Body.condition.score..1.5.,EctoParasitScore = Ectoparasite.score..1.5.,
                    CulmenLmm = Culmen.length..mm.,WingChordLmm = Wing.chord.length..mm.,
                    TarsusLmm = Tarsus.length..mm.,TarsusWmm = Tarsus.width..mm.,HabFlockScore = Habituation.score.of.flock..1.5.)

#This dataset has "-999" values which mean they are intentionally left blank,
#Birds with an ID number of "-999" represent sites that didn't catch any birds,
#but had weather data collected. This code deletes those observations.
IbisField <- IbisField[!(IbisField$ID == "-999"),]

####Site
#These site location abbreviations are the same across datasets
IbisField$Site[IbisField$Site == "Busch Wildlife Sanctuary"] <- "BWS"
IbisField$Site[IbisField$Site == "Cat House (Bonnie's, Richard Lane)"] <- "CAT"
IbisField$Site[IbisField$Site == "Dubois Park"] <- "DUP"
IbisField$Site[IbisField$Site == "DuBois Park"] <- "DUP"
IbisField$Site[IbisField$Site == "Dreher Park"] <- "DRP"
IbisField$Site[IbisField$Site == "Fisheating Creek"] <- "FC"
IbisField$Site[IbisField$Site == "Green Cay"] <- "GC"
IbisField$Site[IbisField$Site == "Gaines Park"] <- "GP"
IbisField$Site[IbisField$Site == "Indian Creek"] <- "ICP"
IbisField$Site[IbisField$Site == "Juno Beach"] <- "JB"
IbisField$Site[IbisField$Site == "Juno Beach "] <- "JB"
IbisField$Site[IbisField$Site == "J.W. Corbett Wildlife Management Area"] <- "JWC"
IbisField$Site[IbisField$Site == "Kitching Creek"] <- "KC"
IbisField$Site[IbisField$Site == "Lion Country Safari"] <- "LCS"
IbisField$Site[IbisField$Site == "Lake Worth"] <- "LW"
IbisField$Site[IbisField$Site == "Loxahatchee Slough"] <- "LOXS"
IbisField$Site[IbisField$Site == "Loxahatchee Wildlife Refuge"] <- "LOXWR"
IbisField$Site[IbisField$Site == "Loxahatchee/LILA Cell B2"] <- "LOXB2"
IbisField$Site[IbisField$Site == "Loxahatchee NE "] <- "LOXNE"
IbisField$Site[IbisField$Site == "Palm Beach Zoo"] <- "WPBZ"
IbisField$Site[IbisField$Site == "Royal Palm"] <- "RP"
IbisField$Site[IbisField$Site == "CROW Wildlife Rehabilitation Center"] <- "CROW"
IbisField$Site[IbisField$Site == "Solid Waste Authority"] <- "SWA"
IbisField$Site[IbisField$Site == "TetraTech"] <- "TT"

####Sex Column
#This replaces the -999 and FAIL values with NA. The IbisBlood datasheets carry more sex data for the birds
IbisField$Sex[IbisField$Sex == "FAIL"] <- NA

####Age Column
IbisField$Age[IbisField$Age == "Adult"] <- "A"
IbisField$Age[IbisField$Age == "3rd year"] <- "J"
IbisField$Age[IbisField$Age == "2nd year"] <- "J"
IbisField$Age[IbisField$Age == "1st year"] <- "J"
IbisField$Age[IbisField$Age == "NA"] <- NA
IbisField$Age[IbisField$Age == "Adult "] <- "A"
IbisField$Age[IbisField$Age == "2nd year "] <- "J"
IbisField$Age[IbisField$Age == "Juvenile"] <- "J"
IbisField$Age[IbisField$Age == "2nd Year"] <- "J"

#Replacing all -999 values in the dataset with NA
IbisField[IbisField == -999] <- NA

#Replacing NA text strings with regular NA values in multiple columns
IbisField$BodyCondScore[IbisField$BodyCondScore == "NA"] <- NA
IbisField$HabFlockScore[IbisField$HabFlockScore == "NA"] <- NA
IbisField$CulmenLmm[IbisField$CulmenLmm == "NA"] <- NA
IbisField$WingChordLmm[IbisField$WingChordLmm == "NA"] <- NA
IbisField$TarsusLmm[IbisField$TarsusLmm == "NA"] <- NA
IbisField$TarsusWmm[IbisField$TarsusWmm == "NA"] <- NA

#--------------------------------------------------------------------------------------#

###############################Saving the Cleaned Dataset###############################

save_data_location3 <- here::here("processed_data", "processedIbisFielddata.rds")
saveRDS(IbisField, file = save_data_location3)
```

This cleaning code section will clean, process, and save the Ibis urbanization dataset. This section is contained in the "ProcessingScriptIbisUrban.R" file.

```{r}

##############################Cleaning the rdUrbanIbis data###########################
#Selects only the columns of interest in the dataset
IbisUrban <- select(rdUrbanIbis, Name:Serotype)

#Renames the columns of interest
IbisUrban <- rename(IbisUrban,ID = Name,Date = Collection.Date,HgPPM = mg.kg.Hg..PPM.,
                    Site = Site.Name,UrbanPercent = Site...Urbanized)

#Change the site names to their abbreviations
IbisUrban$Site[IbisUrban$Site == "Juno Beach"] <- "JB"
IbisUrban$Site[IbisUrban$Site == "Indian Creek"] <- "ICP"
IbisUrban$Site[IbisUrban$Site == "Dubois Park"] <- "DUP"
IbisUrban$Site[IbisUrban$Site == "Dreher Park"] <- "DRP"
IbisUrban$Site[IbisUrban$Site == "Lion Country Safari"] <- "LCS"
IbisUrban$Site[IbisUrban$Site == "Loxahatchee Wildlife Refuge"] <- "LOXWR"
IbisUrban$Site[IbisUrban$Site == "Solid Waste Authority "] <- "SWA"
IbisUrban$Site[IbisUrban$Site == "Gaines Park"] <- "GP"
IbisUrban$Site[IbisUrban$Site == "Kitching Creek"] <- "KC"
IbisUrban$Site[IbisUrban$Site == "Kitching Creek "] <- "KC"
IbisUrban$Site[IbisUrban$Site == "Loxahatchee NE"] <- "LOXNE"
IbisUrban$Site[IbisUrban$Site == "TetraTech"] <- "TT"
IbisUrban$Site[IbisUrban$Site == "Loxahatchee NE "] <- "LOXNE"
IbisUrban$Site[IbisUrban$Site == "Green Cay"] <- "GC"
IbisUrban$Site[IbisUrban$Site == "J.W. Corbett Wildlife Management Area"] <- "JWC"
IbisUrban$Site[IbisUrban$Site == "Loxahatchee Slough"] <- "LOXS"
IbisUrban$Site[IbisUrban$Site == "Lake Worth"] <- "LW"

#--------------------------------------------------------------------------------------#

################################Saving the UrbanIbis Data#############################
#Saving the Ibis Blood 15-17 dataset as an RDS in the processed data folder
save_data_location3 <- here::here("processed_data", "processedIbisUrban.rds")
saveRDS(IbisUrban, file = save_data_location3)

#--------------------------------------------------------------------------------------#
```

This last cleaning section combines the data sheets into one to use for analysis. The urbanization gradient has not been created as of yet, but a meeting this month with those involved with the project will help shed some light on the exact specifics for analysis. This code is located in the "ExploratoryAnalysisScript.R" file.

```{r}
###############################Completing Combination Data Frames#########################
#Outer joining IbisBlood10_14 and IbisBlood15_17, keeping all values of the 10_14 dataset
IbisBlood10_17 <- merge(IbisBlood10_14, IbisBlood15_17, by = c("ID", "Site", "Date", "Sex", "Age",
                                                              "TotalRBC", "HaeParasit"), all=TRUE)
#There is no overlap in the ID numbers between the two, so no recapture is assumed.

#This is a left join, this data frame below only contains birds that were in the IbisBlood10_17
#dataset (all with parasitemia data), with additional information added by the IbisField data. Ibis
#from the Field data that weren't tested for parasitemia or weren't included on Blood dataset
#are not included in this dataframe

#IbisBlood10_17(with Field)_(Parasitemia data)
IbisC <- left_join(IbisBlood10_17, IbisField, by = c("ID", "Site", "Date", "Sex", "Age", "Season", "HabType",
                                                         "IbisNum", "BirdMassG", "BodyCondScore", "EctoParasitScore",
                                                         "CulmenLmm", "WingChordLmm", "TarsusLmm", "TarsusWmm"))

#This section classifies the habitat type for the sites without a habitat type.
IbisC <- IbisC %>%
        mutate(HabType = ifelse(Site == "ICP", "Urban", HabType)) %>%
        mutate(HabType = ifelse(Site == "JB", "Urban", HabType)) %>%
        mutate(HabType = ifelse(Site == "RP", "Urban", HabType)) %>%
        mutate(HabType = ifelse(Site == "LCS", "Urban", HabType)) %>%
        mutate(HabType = ifelse(Site == "LW", "Urban", HabType)) %>%
        mutate(HabType = ifelse(Site == "SWA", "Urban", HabType)) %>%
        mutate(HabType = ifelse(Site == "E", "Natural", HabType)) %>%
        mutate(HabType = ifelse(Site == "WPBZ", "Urban", HabType)) %>%
        mutate(HabType = ifelse(Site == "PO", "Urban", HabType)) %>%
        mutate(HabType = ifelse(Site == "PP", "Urban", HabType)) %>%
        mutate(HabType = ifelse(Site == "DRP", "Urban", HabType)) %>%
        mutate(HabType = ifelse(Site == "SM", "Urban", HabType))

##This section is adding in the seasons for observations that have collection dates but no season.

#Creating intervals for adding seasons based on the date column
int1 <- interval(ymd("2012-12-01"), ymd("2012-12-31"))
int2 <- interval(ymd("2013-06-01"), ymd("2013-08-31"))
int3 <- interval(ymd("2013-12-01"), ymd("2013-12-31"))
int4 <- interval(ymd("2014-02-01"), ymd("2014-03-31"))

#Adding seasons to the ibis without them stated based on the date the sameple was collected
IbisC <- IbisC %>%
         mutate(Season = ifelse(Date %within% int1, "Fall 2012", Season)) %>%
         mutate(Season = ifelse(Date %within% int2, "Summer 2013", Season)) %>%
         mutate(Season = ifelse(Date %within% int3, "Fall 2013", Season)) %>%
         mutate(Season = ifelse(Date %within% int4, "Spring 2014", Season))

#Adding a column for HaeParasit Presence/Absence data
IbisC <- IbisC %>%
    mutate(HaeParasitPA = as.integer(HaeParasit > 0 & !is.na(HaeParasit)))

#Adding a column for transformed HaeParasit data on a logarithm base 10 scale,
#then removing the -Inf values due to there being no parasitemia in those birds. This column
#then only contains logarithm transformed parasitemia data for birds with a positive value
#of parasitemia.
IbisC <- IbisC %>%
    mutate(HaeParasitLog10 = log10(HaeParasit)) %>%
    mutate(HaeParasitLog10 = ifelse(HaeParasitLog10 == "-Inf", NA, HaeParasitLog10))

#Changing the BodyCondScore to an ordinal scale
IbisC <- IbisC %>%
      mutate(BodyCondScore = recode_factor(BodyCondScore, '1' = 'Emaciated',
                                           '2' = 'Underweight', '3' = 'Ideal',
                                           '4' = 'Overweight', '5' = 'Obese',
                                           .ordered = TRUE))

#Changing the class of some variables for analysis. From character to numeric.
C2N <- c(16,17,18,21,22,23,24)

IbisC[ , C2N] <- apply(IbisC[ , C2N], 2,
                       function(x) as.numeric(as.character(x)))

```

This concludes the data cleaning section, we are now ready for the exploratory analysis!

## Exploratory analysis

>>>>

---
title: "Prevalence of *Haemoproteus plataleae* in American White Ibis from Southern Florida"
subtitle: ""
author: Ryan Grunert
date: "`r Sys.Date()`"
site: bookdown::bookdown_site
documentclass: article
output:
  bookdown::word_document2: 
    toc: false
    number_sections: true
  bookdown::html_document2: 
    toc: false
bibliography: ../references.bib
csl: ../apa.csl
---
```{r}
#Loading required packages #for data processing
library(here) #to set paths
library(tidyverse) #all required data manipulation packages
library(xlsx) #to read in excel spreadsheets
library(janitor) #To assist in cleaning up dates and tidy data
library(lubridate) #To extract times from raw data
```

@@

## General Background Information
The data this analysis concerns is American White Ibis (*Eudocimus albus*) blood parasite data that was collected from 2010-2017. The parasite the data focuses on is the protozoan *Haemoproteus*, which causes a chronic blood infection in White Ibises and many other birds. *Haemoproteus* is a vector-borne disease, it is transmitted by *Culicoides* biting midges when they feed on the birds and causes long-lasting parasitemia. Adult healthy birds tolerate haemosporidian infections well, but young birds and birds outside their normal range that are naive to the pathogen are more likely to develop clinical disease. Haemosporidian infections usually cause long-lasting chronic infections in natural hosts, and can cause reduced reproductive success, host fitness, and increased stress and disease susceptibility.

The purpose of this analysis is to determine whether the presence of *Haemoproteus* was increasing over time in Ibis in southern Florida, and if so, determine any demographic trends among the Ibis or temporal or spatial trends among the system at large. Here we investigate whether age, sex, sampling site, capture season, date of capture, body condition, and/or mass of captured Ibis make significant differences on the overall presence of *Haemoproteus* in the birds.

# Methods

## Data Collection

The data was collected by various undergraduate and graduate students of Dr. Sonia Hernandez and Dr. Michael Yabsley at the University of Georgia. Wild American White Ibis were captured during multiple sampling seasons between 2010-2017 in various urban and natural habitat sites from southern Florida. Data was recorded in excel sheets that included information about the sampling sites (habitat type, location, etc.) and various characteristics about the captured Ibis (weight, sex, age, etc.). Ibis were bled and lab diagnostics were performed at the Southeastern Cooperative Wildlife Disease Study (SCWDS) at UGA in order to determine the presence of *Haemoproteus* parasitemia, and if possible, amount of parasitemia in the blood.

## Data Cleaning

The data was organized in 2 excel sheets; one containing all the lab diagnostic data concerning the sampled Ibis and the other containing all the descriptive, demographic, and geographic data about the sampled Ibis. Both excel files were combined while removing uninformative and duplicate variables in both. Variables of importance were standardized once both datasheets were combined. If sampling season was not recorded but sampling date was for an Ibis, season was determined from the date and added. I dichotomized the recorded amount of *Haemoproteus* parasitemia to either "1" (presence) or "0" (absence) by whether or not the Ibis had a parasitemia value of above zero. Locations that did not have presence of *Haemoproteus* and had less than 5 observations with little demographic information were removed from analysis.

## Data Analysis

Univariate relationships between the presence of *Haemoproteus* parasitemia and each predictor were assessed using logistic regression treating *Haemoproteus* parasitemia presence as a categorical outcome. Multivariate classification was also performed, determination of variables to include was performed by sequential forward floating selection. In addition, a Least Absolute Shrinkage and Selection Operator (LASSO) model, a decision tree, and a random forest model were all conducted with *Haemoproteus* parasitemia presence as the outcome variable. The goal of all models was maximization of accuracy and roc_auc on test data using a 10 fold cross-validation (80/20 split, 5 times repeated).

## Study Population

During all of the sampling seasons, 668 Ibis were sampled. Among those sampled, 355 Ibis had a positive *Haemoproteus* parasitemia value. 20 Ibis were removed from the analysis because they did not have a positive *Haemoproteus* parasitemia value and the site from which they were captured had less than or equal to 5 sampled Ibis and had no presence of infected Ibis.

## Exploratory Analysis
